---
title: "ggplot basics "
author: "Ben Bolker"
date: "29 April 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* newest R graphics framework
* based on Wilkinson's ''Grammar of Graphics''
* documented in a book (see below) and on a [web site](http://had.co.nz/ggplot2), as well as an active [mailing list](http://groups.google.com/group/ggplot2)
* explicit mapping from variables to ''aesthetics'': x, y, colour, size, shape
* implements *faceting*
* some data summaries etc. built in
* easier to overlay multiple data sets, data summaries, model predictions etc.
* no 3D plots
* rendering can be slow
* `gridExtra`, `directlabels` packages may be handy; also [ggplot extensions](http://www.ggplot2-exts.org/ggalt.html)

## ggplot intro

mappings + geoms

### Data

Specified explicitly as part of a `ggplot` call:

```{r ggplot1,message=FALSE}
library(mlmRev)
head(Oxboys)
library(ggplot2)
ggplot(Oxboys)
```

But that isn't quite enough: we need to specify a *mapping* between variables (columns in the data set) and *aesthetics* (elements of the graphical display: x-location, y-location, colour, size, shape ...)

```{r ggplot2}
ggplot(Oxboys,aes(x=age,y=height))
```

but (as you can see) that's still not quite enough.  We need to specify some geometric objects (called `geom`s) such as points, lines, etc., that will embody these aesthetics.  The weirdest thing about `ggplot` syntax is that these `geom`s get *added* to the existing `ggplot` object that specifies the data and aesthetics; unless you explicitly specify other aesthetics, they are inherited from the initial `ggplot` call.
```{r ggplot3}
ggplot(Oxboys,aes(x=age,y=height))+geom_point()
```

* many more geoms (lines, bars, etc.)
* summarizers: smooth lines and summaries (`geom_smooth`, `stat_sum`)
* control of scales (e.g. log transforms, colors, etc.)
* faceting (grid and wrap)

See [Karthik's ggplot intro](https://github.com/karthikram/ggplot-lecture) or [my intro for disease ecologists](http://ms.mcmaster.ca/~bolker/eeid/ecology/ggplot.pdf), among many others.

## Techniques for multilevel data

* faceting (= trellis plots = small multiples) vs grouping ("spaghetti plots")
* join data within a group by lines (perhaps thin/transparent)
* can colour lines by group (get legend), but more useful for explanatory than presentation graphics

```{r ggplot_mult,fig.keep="none",message=FALSE}
data("cbpp",package="lme4")
theme_set(theme_bw())
## need 'period' variable to be numeric ...
cbpp2 <- transform(cbpp,
              period=as.numeric(as.character(period)))
## base plot
g0 <- ggplot(cbpp2,aes(period,y=incidence/size))
g1 <- g0 + geom_line(aes(group=herd))
## spaghetti plot
g2 <- g0+geom_line(aes(colour=herd))+
       geom_point(aes(size=size,colour=herd))
## facet instead
(g3 <- g2+facet_wrap(~herd))
## order by average prop. incidence
cbpp3 <- transform(cbpp2,herd=reorder(herd,incidence/size))
g2 %+% cbpp3
g3 %+% cbpp3
## also consider colouring by incidence/order ...
```

Makes it fairly easy to do a simple *two-stage* analysis on the fly:
```{r ggplot_mult2,fig.keep="none",warning=FALSE}
g0+geom_point(aes(size=size,colour=herd))+
    geom_smooth(aes(colour=herd,weight=size),
                 method="glm",
                method.args=list(family=binomial))
```

```{r}
library(lme4)
glm(cbind(incidence,size-incidence)~herd)


lmList(incidence/size~period|herd,
       data=subset(cbpp,herd!="8"),
       weights=size,family=binomial)
